---
title: "321 HW 5 starter.rmd"
author: "Alex Radecki"
date: "October 31, 2022"
output: word_document
---

The code chunk below allows files to be knitted even if there are programming errors.

```{r set_options, message=FALSE}
knitr::opts_chunk$set(error = TRUE, fig.width = 6, fig.asp = 0.618)
```

The packages we need available for this program are: mosaic, dplyr, broom, and olsrr.  

```{r packages, message=FALSE}
library(mosaic)
library(dplyr)
library(broom)
library(olsrr)
```

We begin by reading in the data set named Coal.  This is a .csv file.  The R dataframe is named Coal.  

```{r}
# read in data file
Coal <- read.csv("/SharedProjects/Gabrosek/STA 321/Coal.csv",header=TRUE)
```


****************************************************************
In this homework we want to build a model to predict whether or not a coal mine pillar will be stable.  The response variable is Stability.  We only consider the following potential predictor variables: Depth, Ratio, and Ratio2. 
****************************************************************

1. The code below does some simple exploratory data analysis of Stability, Depth, Ratio, and Ratio2.

```{r}
tally(~Stability, data = Coal)
favstats(~Depth, data = Coal)
favstats(~Ratio, data = Coal)
favstats(~Ratio2, data = Coal)
```

(a) How many (number, not %) of the coal mine pillars were stable?

Answer: 15

(b) Write a sentence to interpret the mean depth of the pillars.  Include units.

Answer: The average pillar depth (meters) for the dataset is 77.7931 meters

(c) What are the minimum and maximum ratio of width to height?

Answer: Min: 0.57, Max: 8.33

(d) Write a one sentence interpretation of the median ratio of strength to stress.

Answer: 50% of the observations have a strength/stress ratio less than or equal to 1.3, while 50% have a ratio greater than 1.3


2. The code below does some simple exploratory data analysis of Depth, Ratio, and Ratio2.

```{r}
histogram(~Depth, data = Coal, type="count", breaks=seq(0,300,25))
histogram(~Ratio, data = Coal, type="count", breaks=seq(0,9,1))
histogram(~Ratio2, data = Coal, type="count", breaks=seq(0,6,1))
```

(a) True/False: The distribution of Depth is unimodal?

Answer: False, there appears to be a second small peak around depth of 250

(b) True/False: Depth has a peak at 12 meters.

Answer: False, there is a peak at a depth of around 25 meters with 12 observations

(c) Multiple Choice: The distribution of Ratio is: (i) roughly symmetric, (ii) skewed left, or (iii) skewed right.

Answer: iii: skewed right

(d) How many of the pillars have a strength to stress ratio of 1 or less?

Answer: 2: 3-4 and 5-6


3.	The R code below creates a jittered variable named YJitter that adds some random noise to the dummy variable Stability. Then, the code makes separate scatterplots with Depth, Ratio, and Ratio2.

```{r}
# jittered Y
Coal <- Coal %>% mutate(YJitter = Stability + rnorm(mean = 0, sd=.05, n = 29))

# jittered scatterplot with Depth
xyplot(YJitter~Depth, data = Coal, ylab="Stability where 1 = Yes", pch = 20)

# jittered scatterplot with Ratio
xyplot(YJitter~Ratio, data = Coal, ylab="Stability where 1 = Yes", pch = 20)

# jittered scatterplot with Ratio2
xyplot(YJitter~Ratio2, data = Coal, ylab="Stability where 1 = Yes", pch = 20)
```

(a) Write a sentence that explains what relationship, if any, there is between Depth of a pillar and Stability.

Answer: There does not appear to be a relationship between depth and stability

(b) Write a sentence that explains what relationship, if any, there is between Width to Height Ratio of a pillar and Stability.

Answer: There appears to be a slight relationship between width/height ratio and stability, where the mine seams start to become stable as the ratio increases (Starting around a ratio of 2)

(c) Write a sentence that explains what relationship, if any, there is between Strength to Stress Ratio of a pillar and Stability.

Answer: There appears to be a relationship between strength/stress ratio and stability, as low ratios around 1 are not stable, where higher ratios (> 1.5) are stable

(d) Multiple Choice: The variable that appears to have the strongest relationship with Stability is: (i) Depth, (ii) Width to Height Ratio, or (iii) Strength to Stress Ratio.

Answer: iii: Strength to stress ratio


4. The code below fits the model with response Stability and predictor Ratio2.  We call this R Object Model1.

```{r}
# fit model
Model1<- glm(Stability ~ Ratio2, data = Coal, family = "binomial")

# summarize model
summary(Model1)
```

(a) True/False: As the strength to stress ratio of the pillar increases, the chance the pillar is stable decreases.

Answer: False, as the strength to stress ratio increases, the probability the pillar is stable increases

(b) True/False: Using level of significance alpha = 0 .05, strength to stress ratio is a statistically significant predictor of pillar stability.

Answer: True, the p-value for Ratio2 = 0.01174 < 0.05

(c) Write out the regression equation.  Use three-non-zero decimals.

Answer: e^(-7.293 + 5.318X1) / 1 + e^(-7.293 + 5.318X1)

(d) Find the probability that a pillar that has strength to stress ratio 1 will be stable.

Answer: 0.122

(e) Find the probability that a pillar that has strength to stress ratio 1.5 will be stable.

Answer: 0.665

5. The code below fits the model with response Stability and predictor variables Depth, Ratio, and Ratio2.  We call this R Object Model2.

```{r}
# fit model
Model2<- glm(Stability ~ Depth + Ratio + Ratio2, data = Coal, family = "binomial")

# summarize model
summary(Model2)
```

(a) Using level of significance alpha = 0.05, strength to stress ratio is not a statistically significant predictor of pillar stability in a model that includes the predictors Depth and Ratio. How is this possible when Ratio2 was a statistically significant predictor in Model1?

Answer: Just because Ratio2 was a statistically significant predictor in one model does not mean it will be significant in another, as adding/removing predictors in a model will impact all other regression coefficients and p-values, which means some predictors may either become significant or no longer be significant.

