---
title: "hw 1 starter.rmd"
author: "Alex Radecki"
date: "August 30, 2022"
output: word_document
---

This code chunk allows R Markdown to knit your output file even if there are coding errors.

```{r message=FALSE}
knitr::opts_chunk$set(error = TRUE)

```

The packages we need available for this program are: mosaic, dplyr, and olsrr.  

```{r}
library(mosaic)
library(dplyr)
library(olsrr)
```

1.  Do Problem 1.1 on page 18 of the online text.  Type your answers below.  For a categorical (qualitative) variable just give me two of the possible categories.

(a) Answer: Tropical, desert

(b) Answer: Quantitative

(c) Answer: Quantitative

(d) Answer: Asian, Caucasian

(e) Answer: Quantitative

(f) Answer: Quantitative

(g) Answer: Quantitative

(h) Answer: Democrat, Republican


2.  Do Problem 1.3 on page 19 of the online text.  Type your answers below.  Make your explanations very brief.

(a)	Answer: Response: Gasoline consumption of cars; Predictor: Number of cylinders

(b)	Answer: Response: College admission; Predictor: SAT scores, GPA

(c)	Answer: Response: Demand; Predictor: Supply

(d)	Answer: Response: Return on a stock; Predictor: Assets, net sales

(e)	Answer: Response: Time to run race; Predictor: Distance, weather

(f) Answer:	Response: Lung cancer; Predictor: Weight, smoker

(g)	Answer: Response: Height/weight of child; Predictor: Height/weight of parents, sex/age of child



*********************************************************
From this point on you will need to code in R to produce output to answer the questions.
*********************************************************

We begin by reading in the data set named senior swimming.  This is a .csv file.  The R dataframe is named Swim.  Note: The RStudio Server includes a data documentation file named Senior Swimming Data Doc that describes the variables in the data set.

```{r}
# read in data file
Swim <- read.csv("/SharedProjects/Gabrosek/STA 321/senior swimming.csv",header = TRUE)
```


3.	Find numerical summaries for the variables Age and Time. See LM 1  - Basic Numerical Summaries on 321 R how-to sheet.    

```{r}
# find numerical summaries
  favstats(~ Age, data=Swim, na.rm=TRUE)
  favstats(~ Time, data=Swim, na.rm=TRUE)
```

(a)	What is the age of the oldest competitor?  Be sure to give units.

Answer: 94 yrs

(b)	What is the median time?  Be sure to give units.

Answer: 521.4 seconds

(c)	Write a sentence that interprets the median from part (b).

Answer: 50% of all times were less than 521.4 seconds


4.	Make histograms of the variables Age (using start value 50, end 100, and jump 5) and Time (using start value 200, end 1400, and jump 50). See LM 1  - Histogram on 321 R how-to sheet.      

```{r}
# histograms
  histogram(~ Age, data=Swim, type="count", breaks=seq(50, 100, 5))
  histogram(~ Time, data=Swim, type="count", breaks=seq(200, 1400, 50))
```

(a)	Describe the distribution of the variable Age.  Mention peaks, shape, and outliers.

Answer: The graph is unimodal with a peak around 60-65 years and is slightly right skewed with potential outlier(s) around 90+ years

(b)	Describe the distribution of the variable Time.  Mention peaks, shape, and outliers.

Answer: The graph is unimodal with a peak around 450-500 secs and is fairly right skewed with outliers around 1000+ secs


5. Make a scatterplot with X = Age and Y = Time. Add the option cex = 0.5 to the code so that you can see points better.  See LM 1  - Scatterplot on 321 R how-to sheet.   

```{r}
# scatterplot
  xyplot(Time ~ Age, data=Swim, cex=0.5, pch=19)
```

(a)	Is there a relationship between Age and Time?  If yes, what function would best describe that relationship?

Answer: Yes, there appears to be a slightly curvilinear relationship

(b) If there is a relationship describe the strength and direction.

Answer: Moderate strength, positive


6.	Find the linear correlation between X = Age and Y = Time.  Include the option use = "pairwise.complete.obs"  to account for missing values. See LM 1  - Linear Correlation on 321 R how-to sheet. 

```{r}
# find correlation coefficient
  cor(Time ~ Age, data=Swim)
```

(a)	What is the value of linear correlation?  Use proper notation.

Answer: r = 0.6470524

(b)	Describe what the linear correlation tells you about the linear relationship between the variables.

Answer: There is a moderate correlation between time and age

(c) Does the value of the linear correlation suggest that a straight-line is the best function to model the relationship between the variables?

Answer: While r = 0.647 suggests a moderately strong linear relationship, that doesn't mean that a stright line is the best fit  
 

7.	What is the straight-line model between X = Age and Y = Time?  Note: I want the model, not the regression equation that has estimates for the β parameters.  Be sure to tell me what the X and Y variables are.

Answer: Y = B0 + B1X + e where Y = Time and X = Age


8. Fit the simple linear regression model and save the results to an R Object named Model1. See LM 1  - Linear Regression on 321 R how-to sheet.  

```{r}
# find least-squares line
  Model1 <- lm(Time ~ Age, data=Swim)
# Use the summary( ) function to get the output from Model1.
  summary(Model1)
```

(a)	Write out the least-squares regression equation. 

Answer: Y^ = -205.7045 + 11.4434X

(b)	Write a one sentence interpretation of the slope in the context of the problem.

Answer: For every 1 year increase in age, time increases by 11.4434 seconds

(c)	What proportion of the variation in Time is explained by the least-squares line?

Answer: 41.87%

(d)	Determine whether there is a statistically significant linear relationship between Age and Time.  Do the following: (i) write out Ho and Ha in notation, (ii) state the value of the test statistic and its degrees of freedom, (iii) state the p-value, (iv) decide whether to reject Ho, and (v) write a couple sentences interpreting your decision referencing the research question.

Answer: 
i) H0: B1 = 0, Ha: B1 != 0
ii) Test statistic = 20.137, df = 563
iii) p-value: < 2.2e-16
iv) p-value < 0.05, we reject the null hypothesis
v) We have sufficient evidence that there is a statistically significant relationship between age and time at the 5% significance level


9.	Make a scatterplot with X = Age and Y = Time that includes the regression line.  Add the option cex = 0.5 to the code so that you can see points better. See LM 1  - Scatterplot with Regression Line on 321 R how-to sheet. 

```{r}
# scatterplot
  xyplot(Time ~ Age, data=Swim, cex=0.5, pch=19, type=c("p", "r"))
```

10.  Find a 96% confidence interval for the slope.  See LM 1  - Confidence Interval on Slope on 321 R how-to sheet.   

```{r}
# C.I. on slope
  confint(Model1, "Age", level=0.96)
```

(a)	What are the lower limit and the upper limit of the C.I.?

Answer: Lower limit: 10.27359, Upper Limit: 12.61327

(b)	Write a one sentence interpretation of the C.I.

Answer: We are 96% confident that the mean change in time (seconds) is between 10.27359 and 12.61327 for a one unit increase in age (years)


11.	Do the following intervals:  See LM 1  - Intervals on Y Given X on 321 R how-to sheet.  

```{r, Q11}
# Find a 96% confidence interval for the mean Time where the Age is 60 years.  
  Conf1 <- predict.lm(Model1, interval="confidence", level=0.96)
# Find a 96% prediction interval for the mean Time where the Age is 60 years.
  Pred1 <- predict.lm(Model1, interval="prediction", level=0.96)
# Observation 20 is for a player with Age = 60 years.  Type Conf1[20,] below to find the confidence interval results.  Type Pred1[20,] below to find the prediction interval results.
  Conf1[20,]
  Pred1[20,]
```

(a)	What is the point estimate for the two intervals?

Answer: 480.9012

(b)	By hand, show how the point estimate was calculated.

Answer: (lwr + upr) / 2

(c)	Write a sentence that interprets the C.I. for the mean response.

Answer: We are 96% confident that the mean time (seconds) where age is 60 years is between 466.8893 and 494.9131

(d)	Write a sentence that interprets the P.I. for an individual response.

Answer: We are 96% confident that the next observed time (seconds) where the age is 60 years is between 214.6989 and 747.1035 
 


**************************************************************************
The code below creates a variable named Age2 that is the Age squared.  The variable is added to the R dataframe Swim.
**************************************************************************

```{r}
Swim <- Swim %>% mutate(Age2 = Age**2)
```

12.	Fit the quadratic linear regression model and save the results to an R Object named Model2.  See LM 2  - Fitting the MLR Model on 321 R how-to sheet.  

```{r}
# find regression equation
  Model2 <- lm(Time ~ Age + Age2, data=Swim)
# Use the summary( ) function to get the output from Model2.
  summary(Model2)
```

(a) What is the quadratic model?  Note: I want the model, not the regression equation that has estimates for the β parameters.  Be sure to tell me what the X and Y variables are.

Answer: Y = B0 + B1X + B2X^2 + e

(b)	Write out the least-squares regression equation. 

Answer: Y = 1008.63498 - 24.61503X + 0.26237X^2

(c)	Is the Age^2 term needed in the model?  Why or why not?

Answer: Yes, the p-value for the age squared term is statistically significant, so we will keep it in the model


 


